Large squeezing in Bose-Einstein condensates with large numbers of atoms

### 2013-07-18 large_number_squeezing_with_atoms.tex
Reworked \pi pulse section.

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\sinc}{sinc}
\newcommand{\abs}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\expect}[1]{\ensuremath{\left<#1\right>}}
\newcommand{\ket}[1]{\ensuremath{\left|#1\right>}}
\renewcommand{\Re}{\operatorname{Re}}
\renewcommand{\Im}{\operatorname{Im}}

@@

The zero-dimensional, two-mode model described in the previous section shows that arbitrarily high squeezing can be achieved using the intrinsic nonlinearities in a BEC, but it ignores the existence of multiple spatial modes.  There are three problems caused by the existence of multiple modes, and these will limit the achievable squeezing.

@@

We might speculate that the effects of these first two problems would be lessened by trap geometries that lead to near-constant density profiles for the BEC.  This should enable higher squeezing.  We cannot make strong conclusions without also considering the effects of coupling between spatial modes, however, which leads to the third potential problem: coupling between modes is inevitable in the presence of a nonlinearity, meaning that the modes cannot be analyzed independently. Coupling between modes will disturb mode-matching, as well as mixing fields with different phase evolution, which can rapidly destroy squeezing.

@@

One way to avoid any requirement to engineer and maintain specific atomic modes is to post-select a spatial mode.  Rather than considering the variance in the total number of atoms in each internal state, it may be possible to consider only the statistics in a spatially filtered area of the atomic cloud.  For example, suppose the atoms were in a Gaussian atomic distribution, but we consider a mode that only includes the atoms within $p$ standard deviations of the centre. For low $p$, this mode shape is considerably closer to the constant density case than a standard Gaussian. The equivalence between the single mode case and this mode is given by
\begin{equation}
\hbar\chi = U \frac{{\mathrm{Erf}} (\sqrt{2p})^3}{{\mathrm{Erf}}(p)^6} \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}}.
\end{equation}
Such spatial filtering can easily produce higher squeezing, but there is the obvious disadvantage that it  significantly reduces the effective number of atoms.  This is a serious issue for most applications of squeezed sources, such as interferometry, where the signal to noise scales with the square root of the flux.  We will focus on methods for producing squeezing with large atomic number.

\section{Using a $\pi$-pulse to improve mode-matching} \label{sec:pipulse}
Modal mismatch is generated by the different atomic potentials seen by the two internal states.  A straightforward method of minimising this effect is to use a $\pi$-pulse to swap the populations of the internal states after half of the evolution. As can be seen by Eqs~(\ref{eqa1}) and (\ref{eqb1}), such a pulse also leaves the quantum statistics of each state unchanged. In the limit where each state remains in a single mode, and exactly half of the atoms are in each state, this technique ensures that both modes have identical mode shapes after the hold time. In practice this will never work perfectly --- due to the density distributions of $\psi_a(\mathbf{r})$ and $\psi_a(\mathbf{r})$ not reflecting the ground state, resulting in non-identical environments for the two modes, some modal dynamics will occur. Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement. More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.

>>>>

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\sinc}{sinc}
\DeclareMathOperator{\erf}{Erf}
\newcommand{\abs}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\expect}[1]{\ensuremath{\left<#1\right>}}
\newcommand{\ket}[1]{\ensuremath{\left|#1\right>}}
\renewcommand{\Re}{\operatorname{Re}}
\renewcommand{\Im}{\operatorname{Im}}

@@

The zero-dimensional, two-mode model described in the previous section shows that arbitrarily good squeezing can be achieved using the intrinsic nonlinearities in a BEC, but it ignores the existence of multiple spatial modes.  There are three problems caused by the existence of multiple modes, and these will limit the achievable squeezing.

@@

We might speculate that the effects of these first two problems would be lessened by trap geometries that lead to near-constant density profiles for the BEC.  This should enable higher squeezing.  We cannot make strong conclusions without also considering the effects of coupling between spatial modes, however, which leads to the third potential problem: coupling between modes is inevitable in the presence of a nonlinearity, meaning that the modes cannot be analysed independently. Coupling between modes will disturb mode-matching, as well as mixing fields with different phase evolution, which can rapidly destroy squeezing.

@@

One way to avoid any requirement to engineer and maintain specific atomic modes is to post-select a spatial mode.  Rather than considering the variance in the total number of atoms in each internal state, it may be possible to consider only the statistics in a spatially filtered area of the atomic cloud.  For example, suppose the atoms were in a Gaussian atomic distribution, but we consider a mode that only includes the atoms within $p$ standard deviations of the centre. For low $p$, this mode shape is considerably closer to the constant density case than a standard Gaussian. The equivalence between the single mode case and this mode is given by
\begin{equation}
\hbar\chi = U \frac{\erf (\sqrt{2p})^3}{\erf(p)^6} \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}}.
\end{equation}
Such spatial filtering can easily produce higher squeezing, but there is the obvious disadvantage that it  significantly reduces the effective number of atoms.  This is a serious issue for most applications of squeezed sources, such as interferometry, where the signal to noise scales with the square root of the flux.  We will focus on methods for producing squeezing with large atomic number.

\section{Using a $\pi$-pulse to improve mode-matching} \label{sec:pipulse}

Modal mismatch is generated by the different atomic potentials seen by the two internal states.  A straightforward method of minimising this effect in the case that modes $\ket{a}$ and $\ket{b}$ are equally occupied after the initial coupling pulse is to apply a $\pi$-pulse to swap the populations of the internal states after half of the evolution.  As can be seen by Eqs.~\eqref{eqa1} and \eqref{eqb1}, such a pulse also leaves the quantum statistics of each state unchanged.  The idea of this scheme is to make the effective potential seen by the atoms in state $\ket{a}$ during the second half of the hold time very similar to that seen by the atoms that were in state $\ket{a}$ during the first half of the hold time.  In the limit of short hold times and for equal occupations of the two modes, this technique reduces the difference in the mode shapes after the hold time.  In practice this will never work perfectly --- due to the density distributions of $\psi_a(\mathbf{r})$ and $\psi_b(\mathbf{r})$ not being in the ground state, their mode shapes will have changed by the time the $\pi$-pulse is applied, and so the spatial dynamics in the second half of the hold time will not be quite the same as those during the first half.  Nonetheless, in many regimes, such a $\pi$-pulse can produce some improvement.  More importantly, as this scheme accepts that the two modes will change shape, and attempts to make them change in a similar fashion, it allows the possibility of much longer hold times. That is, if we are no longer constrained to hold times short enough such that the mode shapes of the atoms in states $|a \rangle$ and $|b \rangle$ do not change significantly, much more squeezing can be extracted from the system.
